### 2021-04-04 newvariant.tex
focuse on S5 and 6

Since the emergence of the new SARS-CoV-2 variant of concern (VoC), estimating its epidemic strength and speed remains one of the key questions in controlling its spread \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease \citep{anderson1991infectious}.
Epidemic speed, characterized by the growth rate $r$ provides information about the population-level time scale of the epidemic and, potentially provides information on when the new variant will replace the existing variant as the dominant strain.
Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.

@@

In addition, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
This implies that the mean generation interval of the new variant is likely to be longer than that of the wild type.
Other studies have considered the possibility that the faster growth rate of the new variant may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
If the generation-interval distribution of the new variant is different from the generation-interval distribution of the wild type, current estimates of the strength of the new variants may be biased.
More information is clearly needed, however. 
And the story may be complicated: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.

@@

When the assumed distribution matches the true distributions, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}F,I).
In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Rv(t)$ during the growth phase and under-estimate (or over-estimate) $ $ during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I).

We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ (\fref{Rtbias}C,F,I).
When generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L).
In addition, the intercept (if we were to perform regression) need not be at zero due to biases in the slope.
Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.

@@

While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant, \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine its spread and control.
In particular, at any given point in epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
As a baseline scenario, we assume $\rw=0\,\textrm{/days}$ and $\delta=0.1\,\textrm{/days}$ (and therefore $\rv=0\,\textrm{/days}$).
In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.

Here, we consider two types of interventions:
a strength-like intervention which affects the kernel by a constant amount $\theta$ regardless of when they were infected ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$) and a speed-like intervention which affect the kernel at a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
In this case, it is clear that we can control the spread of the variant when either $\theta > \Rv$ or $\phi > \rv$---these two quantities ($\theta$ and $\phi$) are also referred to as strength and speed of intervention, respectively.
In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention, both of which provide equivalent measures for effectiveness of intervention, for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than their corresponding pre-intervention strength and speed of the epidemic to control the spread.
Here, we compare how potential differences in generation-interval distributions affect the control of the spread of the new variant using this strength-and-speed framework.

@@

\fref{strengthspeed} compares the effectiveness of strength-based and speed-based interventions across a wide range of assumptions about generation-interval distributions.
We begin by assuming that the variant and the wild type have equal mean generation intervals and model interventions such that they both reduce $\Rv$ to 0.9 after intervention (\fref{strengthspeed}A).
If the mean generation interval of the variant is longer than that of the wild type, then the strength of the variant will be higher;
since the strength of the strength-like intervention is insensitive to the underlying generation-interval distribution, we may fail to control the spread of the new variant.
The strength of the speed-like intervention, however, depends on the new variant:
since speed-like interventions, such as contact tracing, have disproportionate effects on preventing later transmission, they are more effective (therefore stronger) when the mean generation interval of the new variant is longer.

The speed-based paradigm gives same results about the control but provides important insights (\fref{strengthspeed}B).
The observed speed of the variant at any given moment does not depend on our estimates of its mean generation interval.
Therefore, if we can prevent infections at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
On the other hand, differences in mean generation intervals between the variant and the wild type imply different post-intervention epidemic speed (therefore, intervention speed) under a constant-strength intervention, explaining why it might not be effective if the variant has longer generation intervals.

@@

Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
We define relative strength as the ratio between reproduction numbers ($\rho=\Rv/\Rw$) and relative speed as the difference between growth rates ($\delta=\rv-\rw$) of the variant and the wild type.
We show that longer generation intervals generally lead to higher relative strength and lower relative speed, given that the other quantities are fixed, when the incidence of variant is increasing (i.e., $\Rv > 1$ and $\rv > 0$).
We also show that propagating uncertainty in possible differences in mean generation intervals alone can result in considerable variation in the estimates of relative strength $\rho$.
We illustrate how these biases might be assessed in practice and further discuss how differences in generation intervals might lead to different conclusions about the effectiveness of interventions:
if the variant has longer or shorter generation intervals than the wild type, speed- or strength-like interventions, respectively, will be more effective and therefore should be prioritized.

>>>>

Since the emergence of SARS-CoV-2 variants of concern (VoCs), estimating variant epidemic strength and speed remains one of the key questions in controlling its spread \citep{switzerland2021variant, davies2021estimated, di2021impact, leung2021early, volz2021transmission,zhao2021}.
Epidemic strength, characterized by the reproduction number $\RR$, provides information about the final size of the epidemic as well as the amount of intervention required to eliminate the disease \citep{anderson1991infectious}.
Epidemic speed, characterized by the growth rate $r$ provides information about the population-level time scale of the epidemic and, potentially provides information on when a new variant will replace the existing variant as the dominant strain.
Strength and speed are linked by generation intervals---defined as time between infection and transmission---which characterizes individual-level time scale of the epidemic \citep{roberts2007model,svensson2007note,wallinga2007generation}.

@@

\jd{I have started editing to make us a little less focused on B.1.1.7; we're not completely clear whether this is mostly a paper about B1, or whether it's just an example. I think we should focus less, and think also about new variants that might be coming. I stopped here, because this \P\ seems pretty clearly B1-focused and that part seems fine. Something to keep in mind.}
In addition, most studies have assumed that the generation-interval distributions do not differ between the existing strain and the variant, but recent evidence suggests that the new variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
This implies that the mean generation interval of the new variant is likely to be longer than that of the wild type.
Other studies have considered the possibility that the faster growth rate of the new variant may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
In general, if the generation-interval distribution of a new variant is different from the generation-interval distribution of the wild type, estimates of variant strength may be biased.
And the story may be complicated: generation intervals depend on many other biological and behavioral factors, including for example self-isolation after symptom onset leading to shorter generation intervals.

@@

\jd{It would be nice to have the R axes logged.}

When the assumed distribution matches the true distributions, the estimated reproduction numbers match the true values (\fref{Rtbias}A), and therefore the ratio remains constant (\fref{Rtbias}B).
However, if the generation-interval distribution of the variant is different from the assumed distribution, the ratio changes over time (\fref{Rtbias}F,I).
In particular, if the true generation intervals have a shorter (or longer) mean than the assumed distribution, we over-estimate (or under-estimate) $\Rv(t)$ during the growth phase and under-estimate (or over-estimate) during the decay phase (\fref{Rtbias}D,G), which further translates to biases in the estimated relative strength (\fref{Rtbias}F,I). \jd{Not quite following the logic of how you're linking to the pictures. I note that panels C, E and H are not mentioned.}

We can also assess these biases by plotting the estimated strength of the variant $\hat{\RR}_{\textrm{var}}(t)$ against the estimated strength of the wild type $\hat{\RR}_{\textrm{wt}}(t)$ (\fref{Rtbias}C,F,I).
When generation-interval distributions are correctly specified, we obtain a straight line with a slope of $\rho$ and intercept at zero (\fref{Rtbias}C).
When the assumed mean generation interval is longer (or shorter) than the that of the variant, we over-estimate (or under-estimate) the slope (\fref{Rtbias}H,L).
In addition, the intercept (if we were to perform regression) need not be at zero due to biases in the slope.
Smooth changes in $\RR$ further generate circular patterns but biases in slopes are clear.
\jd{Maybe you didn't finish editing here? You've now mentioned F and three times, but still not E or H. And I can't find L despite repeatedly updating. It also doesn't quite seem to match the caption.}

@@

While relative speed $\delta$ and strength $\rho$ are useful for characterizing the spread of the variant, \emph{absolute} speed $\rv$ and strength $\Rv$ of the variant are the main variables that determine its spread and control.
In particular, at any given point in epidemic, we can measure the speed of the variant $\rv$ (or infer $\rv$ from $\rw$ and $\delta$) and ask how much more intervention is required to control the spread of both strains (since $\Rv < 1$ implies $\Rw < 1$).
As a baseline scenario, we assume $\rw=0$ and $\delta=0.1\pday$ (and therefore $\rv=0.1\pday$).
\jd{This sentence invites arguments, and we don't need it anyway:} In this case, the wild type is already under control, whereas the variant requires additional intervention measures to stop growing.

We consider two types of intervention:
an intervention of constant ``strength'', reduces transmission by a constant factor $\theta$ regardless of age of infection ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau)/\theta$); and an intervention of constant ``speed'' which reduces transmission after infection by a constant rate $\phi$ ($K_{\mathrm{post}}(\tau) = K_{\mathrm{pre}}(\tau) \exp(-\phi \tau)$).
In this case, we can control the spread of the variant when $\theta > \Rv$ or $\phi > \rv$, respectively.
In general, we can calculate the strength ($\theta = \Ry{pre}/\Ry{post}$) and speed ($\phi = \ry{pre} - \ry{post}$) of intervention for any intervention by comparing epidemiological conditions before and after the intervention \citep{doi:10.1098/rspb.2020.1556}---these quantities, by definition, must be greater than their corresponding pre-intervention strength and speed of the epidemic to control the spread.
Here, we use the strength-and-speed framework to compare how differences in generation-interval distributions would affect the control of the spread of a new variant.

@@

\fref{strengthspeed} compares the effectiveness of strength-based and speed-based interventions across a wide range of assumptions about generation-interval distributions.
We begin by assuming that the variant and the wild type have equal mean generation intervals and model interventions such that they both reduce $\Rv$ to 0.9 after intervention (\fref{strengthspeed}A).
If the mean generation interval of the variant is longer than that of the wild type, then the strength of the variant will be higher;
since the strength of the strength-like intervention is insensitive to the underlying generation-interval distribution, we may fail to control the spread of the new variant.
With speed-like interventions, however, there is a countervailing effect:
since speed-like interventions like contact tracing have disproportionate effects on preventing later transmission, they are more effective (therefore stronger) when the mean generation interval of the new variant is longer.

The speed-based paradigm gives the same results about the control but provides important insights (\fref{strengthspeed}B).
The observed speed of the variant at any given moment does not depend on our estimates of its mean generation interval.
Therefore, if we can reduce infectiousness at a faster rate than the observed speed of spread (i.e., if the speed of intervention is faster than the epidemic speed), we can control the epidemic---this explains why a constant-speed intervention can prevent the spread of the variant regardless of its mean generation interval.
\jd{This s.~feels repetitive, do we need it?:}
On the other hand, differences in mean generation intervals between the variant and the wild type imply different post-intervention epidemic speed (therefore, intervention speed) under a constant-strength intervention, explaining why it might not be effective if the variant has longer generation intervals.

@@

Here, we explored how the generation-interval distribution shapes the link between relative strength and speed of the new variant.
We define relative strength as the ratio between reproduction numbers ($\rho=\Rv/\Rw$) and relative speed as the difference between growth rates ($\delta=\rv-\rw$) of the variant and the wild type.
We show that longer generation intervals generally lead to higher relative strength for a given relative speed (and conversely, lower relative speed for a given relative strength); these can be reversed when incidence is decreasing.e., $\Rv > 1$ and $\rv > 0$).
We also that uncertainty about the relative mean generation interval of the variant has important implications for estimates of relative strength of the variantj.
We illustrate how these biases might be assessed and discuss how differences in generation intervals might lead to different conclusions about the effectiveness of interventions:
if the variant has longer (respectively, shorter) generation intervals than the wild type, speed-like (strength-like) interventions, will be more relatively more effective than naive estimates would suggest.
